---
title: "Cumulative"
author: "Faisal Jan"
format: html
execute: 
  echo: false
---

```{r}
#| message: false
#| warning: false
library(MASS)
library(tidyverse)
library(primer.data)
library(broom)
library(easystats)
library(gt)
library(marginaleffects)

tidy_approval <- readRDS("data/tidy_approval.rds")
```

```{r}
x <- ces |>
  filter(year == 2020) |>
  select(approval, ideology, education) |>
  drop_na() |>
  filter(!ideology %in% "Not Sure") |>
  mutate(ideology = fct_drop(ideology))
```

```{r}
x |>
  ggplot(aes(x = ideology, fill = approval)) +
  geom_bar() +
  labs(
    title = "Presidential Approval By Ideology",
    subtitle = "Most liberals strongly disapprove the 2020 president.",
    x = "Ideology",
    y = "Count",
    caption = "Source: Cooperative Election Study",
    fill = "Approval"
  )
```

People with different ideologies differ in approval ratings for any political positions. Using data from the 2020 Cooperative Election Study, I seek to explore how presidential approval varies across different ideological groups. Some of these approvals are self-reported and by YouGov panelists that can add bias to the data. I model presidential approval as an outcome of the ordinal function of political ideology and education. The predicted probability of a Very Conservative individual with a Post-Grad education Strongly Approving is estimated at 75%, with a 95% confidence interval ranging from 72% to 78%.

$$
\begin{aligned}
\text{ Cumulative Logistic } \\
P(Y \le k) = \frac{\exp(\theta_k - \eta)}{1 + \exp(\theta_k - \eta)}
\end{aligned}
$$

where:
* $Y$ is the ordinal outcome variable (e.g., `approval`).
* $k$ is the category threshold, with $K-1$ unique thresholds to be estimated.
* $\theta_k$ is the intercept for the $k$-th threshold.
* $\eta$ is the linear predictor, which is a function of the independent variables:
    $$
    \eta = \beta_1 X_1 + \beta_2 X_2 + \cdots + \beta_n X_n
    $$

This structure ensures that the probabilities are correctly ordered, as the model estimates an increasing series of thresholds ($\theta_1 < \theta_2 < \cdots < \theta_{K-1}$).

```{r}
fit_approval <- polr(approval ~ ideology + education, data = x)
```

```{r}
#| cache: true
fit_approval |> tidy(conf.int = TRUE)
```

$$
\log\left(\frac{P(\hat{\text{approval}} \le k)}{1 - P(\hat{\text{approval}} \le k)}\right) = \hat{\theta}_k - \left( \begin{aligned}
& 0.615 \cdot \text{ideology}_{\text{Liberal}} \\
& + 2.40 \cdot \text{ideology}_{\text{Moderate}} \\
& + 4.46 \cdot \text{ideology}_{\text{Conservative}} \\
& + 5.45 \cdot \text{ideology}_{\text{Very Conservative}} \\
& - 0.0957 \cdot \text{education}_{\text{High School Grad}} \\
& - 0.311 \cdot \text{education}_{\text{Some College}} \\
& - 0.325 \cdot \text{education}_{\text{2-Year}} \\
& - 0.521 \cdot \text{education}_{\text{4-Year}} \\
& - 0.727 \cdot \text{education}_{\text{Post-Grad}}
\end{aligned} \right)
$$

Where:
\begin{itemize}
    \item $\hat{\text{approval}}$ represents the fitted ordinal dependent variable.
    \item $k$ is the index for the cumulative category, ranging from $1$ to $4$.
    \item $\text{ideology}_{\text{Level}}$ and $\text{education}_{\text{Level}}$ are dummy variables, taking a value of $1$ if the individual's `ideology` or `education` matches that level, and $0$ otherwise. The implicit reference levels are the unlisted categories for each factor (e.g., the base level of `ideology` and `education`).
\end{itemize}

The estimated thresholds ($\hat{\theta}_k$) are:
\begin{itemize}
    \item $\hat{\theta}_1$ (Strongly Disapprove $| \text{Disapprove / Somewhat Disapprove}$) $= 2.26$
    \item $\hat{\theta}_2$ (Disapprove / Somewhat Disapprove $| \text{Neither Approve nor Disapprove}$) $= 2.73$
    \item $\hat{\theta}_3$ (Neither Approve nor Disapprove $| \text{Approve / Somewhat Approve}$) $= 2.82$
    \item $\hat{\theta}_4$ (Approve / Somewhat Approve $| \text{Strongly Approve}$) $= 3.93$
\end{itemize}

This model estimates the cumulative log-odds of an individual's approval level based on their ideology and education, assuming proportional odds across the categories.

```{r}
tidy_approval |>
  select(term, estimate, conf.low, conf.high) |>
  gt() |>
  tab_caption(
    "Estimated Coefficients and Thresholds for Cumulative Logit Model of Approval"
  ) |>
  cols_label(
    term = "Variable",
    estimate = "Estimate",
    conf.low = "Lower CI",
    conf.high = "Upper CI"
  )
```

```{r}
plot_predictions(
  fit_approval,
  condition = c("ideology", "education"),
  draw = FALSE
) |>
  ggplot(aes(x = ideology, y = estimate, color = group, group = group)) +
  geom_line(linewidth = 1.2, alpha = 0.8) +
  geom_point(size = 3, alpha = 0.8) +
  facet_wrap(~education, scales = "free_y", ncol = 3) + # Increased columns for wider facets
  labs(
    title = "Presidential Approval Probabilities Across Ideology and Education",
    subtitle = "Predicted presidential approval increases with conservative ideology and higher education levels.",
    x = "Ideology",
    y = "Predicted Probability",
    color = "Approval Level"
  ) +
  scale_y_continuous(
    labels = scales::percent_format(),
    breaks = seq(0, 1, by = 0.2)
  ) +
  scale_color_viridis_d(option = "D", begin = 0.1, end = 0.9) +
  theme_minimal(base_size = 12) + # Slightly reduced base font size
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold", size = 14),
    plot.subtitle = element_text(hjust = 1, size = 10),
    axis.title = element_text(face = "bold", size = 12),
    axis.text.x = element_text(angle = 45, hjust = 1, size = 9), # Smaller x-axis labels
    legend.title = element_text(face = "bold", size = 11),
    legend.text = element_text(size = 9),
    legend.position = "bottom", # Move legend to bottom
    legend.box = "horizontal", # Arrange legend items horizontally
    legend.key.size = unit(0.7, "lines"), # Smaller legend key squares
    strip.text = element_text(face = "bold", size = 10), # Smaller facet titles
    panel.grid.minor = element_blank(),
    panel.spacing = unit(0.8, "lines") # Reduced spacing between facets
  )
```
